Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2019-08-14 peakcharacterization_manuscript.Rmd
changing the default chunk options more appropriate for a manuscript
comparing peak ordering

@@

On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation]. These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points). The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb. Our instrument seems to have a mode of 0.5, as shown in Figure XC and XD. Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE. This model seems to fit the known relationship between frequency and M/Z. This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points. M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z. Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

@@

These models are calculated for each scan, and the square root term from all scan level models checked for outliers based on the interquartile ranges across all scans in a sample. When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA. Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.

```{r peak_ordering}
# Figure showing changes in order of peaks
# a - peaks in M/Z compared to frequency scan specific
# b - peaks in M/Z compared to frequency single model

```

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

@@

On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation]. These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points). The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure XB and XC. The Thermo Fusion instrument from which most of our collaborators data has been acquired, at resolution of 400K, seems to have a mode of 0.5, as shown in Figure XC and XD. Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE. This model seems to fit the known relationship between frequency and M/Z. This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points. M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z. Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

@@

Frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample. When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA. Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.

```{r peak_ordering}
drake::loadd(plot_peak_ordering)
plot_peak_ordering
```
